Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-13 linear_adia.tex
updating

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode and represents
the entire disk column moving vertically. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},
and are the ones we find to develop in protoplanetary disks
with a realistic estimate for thermal timescales
(\S\ref{application}).

%To determine the (assumed) small effect of introducing a short thermal
%relaxation, 
We linearize Eq. \ref{nearly_iso_explicit} about the above
solution for $\beta\equiv0$ and write 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta,
\end{align}
with $\delta v_z$ and $\hat{\sigma}$ given by Eq. \ref{pert_basic}. We
may then seek  
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

@@

\begin{align}
  0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
  \left(\hat{\sigma}^2 -1\right)\delta\chi,\notag\\
  0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
    q \khat\right)\delta\chi.
\end{align}

Using the expressions for $\delta\chi$ and $\hat{\sigma}$ we can solve
for $\delta\hat{\sigma}$. This gives 
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
      q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by small finite cooling is ineffective at very large or
very small scales.

@@

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
For ease of analysis, we square Eq. \ref{adia_disp0} to obtain a 
a polynomial in $\hat{\sigma}$,  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ generally
requires a numerical solution. However, simple solutions may be
obtained in limiting cases which we consider below.

@@

\subsection{Upper limit on the thermal relaxation timescale for the
  VSI}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the VSI. We consider wavenumbers such that
$\khat^2\gg M+1$ which, as we will find, correspond to low frequency
modes.

%At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$ is real.  
%This is a pair of simultaneous
%equations for $\hat{\omega}$ and $\beta$. To proceed, we
%consider further limits.

Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} at marginal
stability with $\khat\gg 1$. We consider modes with $M=0$ (the
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$ because $\khat\gg 1$. We
also assume $|q|$ and $\beta$ are $O(1)$ or less. For a thin disk
($\epsilon \ll 1$), we have approximately
\begin{align}
 0 = &M(M+1) + \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2\notag\\
   &-  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and the critical thermal
timescale $\beta$ for marginal stability. If $\beta\ll 1$,
then a simple solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq M+1, 
\end{align}
which is the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Then  
\begin{align}\label{bcrit_gen}
  \beta = \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{(1+M)^{3/2}}{(1+2M)^2} \leq \beta_\mathrm{crit} 
\end{align}
where 
\begin{align}\label{iso_vsi_cond}
  \beta_\mathrm{crit}  \equiv \frac{\epsilon|q|}{\gamma-1}
\end{align}
is the critical thermal timescale for the fundamental mode
($M=0$). However, according to Eq. \ref{bcrit_gen}  the critical
thermal timescale decreases with increasing  $M$. We thus expect
higher order modes to be stabilized more rapidly by finite thermal
relaxation. When $\beta>\beta_\mathrm{crit}$, all modes
(satisfying the above assumptions) are stabilized. We can interpret
$\beta_\mathrm{crit}$ as the thermal timescale below which
perturbations are effectively isothermal to allow the VSI.  We will confirm
Eq. \ref{iso_vsi_cond} numerically in \S\ref{bcrit_num_test}.

(We remark that the critical thermal timescale $\beta_\mathrm{crit}$
for the fundamental mode may be obtained more readily with fewer
assumptions by setting $M=0$ in
Eq. \ref{relax_cond1}---\ref{relax_cond2} from the onset.)

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. This 
conclusion was reached in these studies by considering
vertically-localized disturbances. % , although a definitive timescale
                                % was not presented.  
By contrast, our Eq. \ref{iso_vsi_cond} quantifies this requirement
for vertically global disturbances. We see that in a thin disk, a very
short thermal timescale is required ($\beta_\mathrm{crit}\ll 1$)
because $\beta_\mathrm{crit}\propto \epsilon$. 
This suggests that finite thermal relaxation rapidly stabilizes the
VSI.

>>>>

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to dominate in numerical calculations with 
increasing $\beta$, as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).  %Their importance is highlighted

We linearize Eq. \ref{nearly_iso_explicit} about the above
solution for $\beta\equiv0$ and write 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta,
\end{align}
with $\delta v_z$ and $\hat{\sigma}$ given by Eq. \ref{pert_basic}. We
may then seek  
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

@@

\begin{align}
  0 &= 2a + 2\hat{\sigma}\left(1+\khat^2\right)\delta\hat{\sigma} +
  \left(\hat{\sigma}^2 -1\right)\delta\chi, \\
  0&= 2a\left(1+\ii\epsilon q \khat\right) - \left(\khat^2 - \ii\epsilon
    q \khat\right)\delta\chi.
\end{align}

Using the expressions for $\delta\chi$ and $\hat{\sigma}$ we can solve
for $\delta\hat{\sigma}$. We find imaginary part of $\delta\hat{\sigma}$
is 
% This gives 
% \begin{align}\label{nearly_iso_dsig}
%   \delta \hat{\sigma} =
%   -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
%       q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
% \end{align}
% The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by small finite cooling is ineffective at very large or
very small scales.

@@

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
For ease of analysis, we re-arrange Eq. \ref{adia_disp0} and square it
to obtain a polynomial in $\hat{\sigma}$,  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ is
complicated and generally requires a numerical solution. However,
simple solutions may be obtained in limiting cases which we consider
below.

@@

\subsection{Upper limit on the thermal relaxation timescale for the
  VSI}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta$ that allows the VSI for modes with $\khat^2\gg M+1$.  
% which, as we will find, correspond to low frequency
% modes. 
Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
stability and $\khat^2\gg 1$. We consider modes with $M=0$ (the
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$. We
also assume $|q|$ and $\beta$ are $O(1)$ or smaller. For a thin disk
($\epsilon \ll 1$) we have 
\begin{align}
 0 = &M(M+1)\left(1-\epsilon^2q^2\khat^2\right) \notag\\ 
 &+ \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
   -  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that
$|\epsilon q \khat |\ll 1$ 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq M+1, 
\end{align}
which is the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Then  
\begin{align}\label{bcrit_gen}
  \beta = \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)} \leq \beta_\mathrm{crit} 
\end{align}
where 
\begin{align}\label{iso_vsi_cond}
  \beta_\mathrm{crit}  \equiv \frac{\epsilon|q|}{\gamma-1} 
\end{align}
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without assuming $|\epsilon q \khat|\ll 1$.  We
will confirm Eq. \ref{iso_vsi_cond} numerically in \S\ref{bcrit_num_test}.

Eq. \ref{bcrit_gen} suggests that as the thermal timescale is
increased from zero, higher order modes with $M>0$
are stabilized before the fundamental mode. We indeed observe this
trend in numerical calculations, despite adopting somewhat restrictive
assumptions in order to arrive at the simple expression 
Eq. \ref{bcrit_gen} for $M>0$.

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}, but these
studies consider vertically localized disturbances. Our 
Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
global disturbances (which have small $M$).  We can interpret
$\beta_\mathrm{crit}$ as the thermal timescale below which
perturbations are effectively isothermal and the VSI operates.  We see
that in a thin disk, a very short thermal timescale is required
($\beta_\mathrm{crit}\ll 1$) because $\beta_\mathrm{crit}\propto \epsilon$.

### 2015-03-13 linear_iso.tex
restructured
appendix.tex
updating

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. % We examine isothermal perturbations ($\beta =
% 0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

\subsection{Relation to the vertically global
  shearing box} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent. This
is an attempt to capture the effect of vertical shear in a local model
of accretoin disks. The difference between the radially-local
approximation and the VGSB is that the radial disk structure is
ignored in the latter. That is, the  equilibrium density and pressure
field in the VGSB is homogeneous in the radial ($x$) direction, as in
the standard shearing box.

@@

% but modified by the vertical shear term
% $r\p_z\Omega\delta \delta v_z$.

After setting $\hat{g}_c=0$ in Eq. \ref{lin_mass}, \ref{lin_vx} and
\ref{lin_energy}, we can eliminate the perturbed velocities between
the linearized equations to obtain a pair or ordinary differential
equations for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
    \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
  &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
    \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
  &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
  \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

The VGSB is a self-consistent framework but extrapolating results 
obtained from the VGSB to the global disk has limitations. This is
because within the VGSB framework, $d\Omega/dz$ can be freely
specified. In effect, by adopting the VGSB framework, we are studying
dynamics in the standard shearing box subject to an additional
azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
input parameter. For consistency with the global disk, we choose
$d\Omega/dz$ in the VGSB to be that due to the thermal structure in
the global disk (Eq. \ref{vertical_shear}). However, 
% This is different to the global problem because in
% that case, the vertical shear is a result of the 
the background radial disk structure is ignored elsewhere in the VGSB
(which is done by setting $\hat{g}_c=0$).

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
We find that for perturbations with small radial lengthscales
($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

@@

\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ for $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  % &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
  %     \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
  &+ \left[\frac{\sigma^2}{c_s^2}\left(\frac{k_x^2c_s^2}{\Omega_k^2} +
      \chi\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
  &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
    \left(1-\chi\right) 
  \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks.

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplaneary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion. However, this
approximation will be relaxed in our numerical study.

Note that we can generalize Eq. \ref{vertiso_gov_nondim} non-Keplerian
rotation profiles by redefining $\khat$ as $k_xc_s/\kappa$,
$\hat{\sigma}$ as $\sigma/\kappa$, and  $\epsilon$ as $c_s/r\kappa$.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. % We examine isothermal perturbations ($\beta =
% 0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.  However, all of the 
approximations described in
\S\ref{vgsb_approx}---\S\ref{analytic_relax} are relaxed in our
numerical study.

\subsection{Relation to the vertically global
  shearing box}\label{vgsb_approx} 
We connect the linearized equations 
Eq. \ref{lin_mass}---\ref{lin_energy} to the vertically global  
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is an extension of the standard shearing box \citep{goldreich65} to
allow the shear flow in the shearing box to be height-dependent. 
The difference between the radially local
approximation and the VGSB is that the equilibrium disk
structure is assumed to be homogeneous in the radial ($x$), as in the
standard shearing box.

@@

After setting $\hat{g}_c=0$ in Eq. \ref{lin_mass}, \ref{lin_vx} and
\ref{lin_energy}, we can eliminate variables % eliminate the perturbed velocities between
% the linearized equations 
to obtain a pair or ordinary differential
equations for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
    \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
  &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
    \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
  &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
  \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

Extrapolating results obtained from the VGSB to the global disk has
limitations. This is because within the VGSB framework, $d\Omega/dz$
can be freely specified. %  In effect, by adopting the VGSB framework, we are studying
% dynamics in the standard shearing box subject to an additional
% azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
% input parameter. 
%For consistency with the global disk, 
We choose $d\Omega/dz$ to be that due to the thermal structure in 
the global disk (Eq. \ref{vertical_shear}). However, background radial
gradient terms are ignored elsewhere in the VGSB (which is done by
setting $\hat{g}_c=0$).

%the above inconsistency

We further discuss caveats of the VGSB in Appendix
\ref{global_corr}. We find that for perturbations with small radial
lengthscales ($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$),
which characterizes the VSI, the VGSB framework is adequate for the
purpose of a simple analytical treatment of the VSI.

@@

\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$. The result for $\Gamma=1$ is
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  % &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
  %     \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
  &+ \left[\frac{\sigma^2}{c_s^2}\left(\frac{k_x^2c_s^2}{\Omega_k^2} +
      \chi\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
  &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
    \left(1-\chi\right) 
  \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} =\hat{\omega} +
  \ii\hat{\nu}= \sigma/\Omega_k,
\end{align}
where $\hat{\omega}$ and $\hat{\nu}$ are real. 
Eq. \ref{vertiso_gov} becomes 
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$.%  Eq. \ref{vertiso_gov_nondim} will
% be the basis for our discussion of vertically isothermal disks.

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplaneary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion. % However, this
% approximation will be relaxed in our numerical study.

% Note that we can generalize Eq. \ref{vertiso_gov_nondim} non-Keplerian
% rotation profiles by redefining $\khat$ as $k_xc_s/\kappa$,
% $\hat{\sigma}$ as $\sigma/\kappa$, and  $\epsilon$ as $c_s/r\kappa$.
